---
title: "Life Expectancy by Year"
author: "Mu-Wei Chung"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(plotly)
```

# Life Expectancy Dataset

[Life Expectancy](https://ourworldindata.org/life-expectancy#all-charts) dataset from Our World in Data.

```{r}
le_df <- read_csv("data/life-expectancy.csv")
glimpse(le_df)
```

## Cleaning

```{r}
le_clean <- janitor::clean_names(le_df)
# rename column
le_clean <- le_clean %>% 
  rename(period_life_expect = period_life_expectancy_at_birth_sex_all_age_0)
glimpse(le_clean)
```
## Entity

It include countries, continents, world, groups of different development status and income levels.

```{r}
le_clean %>% pull(entity) %>% unique()
```

## Missing values

```{r}
visdat::vis_miss(le_clean)
```

Which entities are missing codes? 

Continents, Regions by Income and development status

```{r}
le_clean %>% 
  filter(is.na(code)) %>% 
  select(entity) %>% 
  distinct()
```

## World Life Expectancy by Year

```{r}
p <- le_clean %>% 
  filter(entity == "World") %>% 
  ggplot(aes(x = year, y = period_life_expect)) +
  geom_line() +
  labs(title = "World Life Expectancy by Year",
       x = "Year",
       y = "Life Expectancy at Birth (years)")
ggplotly(p)
```

## Looking into different development status regions

```{r}
development_status <- stringr::str_subset(le_clean$entity, regex("devel", ignore_case = TRUE)) %>% unique()

p <- le_clean %>%
  filter(entity %in% development_status) %>%
  ggplot(aes(x = year, 
             y = period_life_expect,
             color = entity)) +
  geom_line() + 
  labs(title = "Life Expectancy by Year",
       x = "Year",
       y = "Life Expectancy at Birth (years)")
ggplotly(p)
```

How many countries have data before 1950? 

```{r}
le_clean %>% drop_na(code) %>%
  filter(year < 1950) %>% 
  pull(entity) %>% 
  unique() %>% 
  length()
```

Only 87 of 238 countries have data before 1950.

```{r}
le_clean %>% drop_na(code) %>%
  filter(year >= 1950) %>% 
  group_by(entity) %>% 
  count() %>% 
  filter(n == 72) %>%
  pull(entity) %>%
  unique() %>%
  length()
  
```

237 of 238 countries have data every year since 1950.

```{r}
le_clean %>% drop_na(code) %>%
  pull(entity) %>%
  unique() %>% 
  length()
```


# Difference in Life Expectancy

Create a new column that calculates the difference in life expectancy from the previous year.

```{r}
le_clean <- le_clean %>% 
  group_by(entity) %>% 
  mutate(diff = c(NA, diff(period_life_expect))) %>%
  ungroup()
```

```{r}
p <- le_clean %>%
  filter(entity %in% development_status) %>%
  drop_na(diff) %>%
  ggplot(aes(x = year, 
             y = diff,
             color = entity)) +
  geom_line() + 
  geom_abline(intercept = 0.243, slope = 0, color = "black", linetype = "dotted") +
  labs(title = "Life Expectancy Diff by Year",
       x = "Year",
       y = "Life Expectancy at Birth (years)")
ggplotly(p)
```

# Time Series Analysis

## Convert dataframes to time series

```{r}
# create a list of 3 dataframes
le_dev_ls <- list()
for (dev in development_status) {
  le_dev_ls[[dev]] <- le_clean %>% 
    filter(entity == dev) %>%
    select(period_life_expect) %>%
    ts(start = 1950, end = 2021)
}
le_dev_ls
```

## Fit best ARIMA model

```{r}
# install.packages("forecast")
library(forecast)
le_dev_opt <- list()
for (dev in development_status) {
  le_dev_opt[[dev]] <- auto.arima(le_dev_ls[[dev]])
}
```

## Forecast the life expectancy for the next 5 years
```{r}
p1 <- le_dev_opt$`Least developed countries` %>% 
  forecast(h = 5) %>% 
  autoplot()
p2 <- le_dev_opt$`More developed regions` %>%
  forecast(h = 5) %>% 
  autoplot()
p3 <- le_dev_opt$`Less developed regions, excluding least developed countries` %>%
  forecast(h = 5) %>% 
  autoplot()
p4 <- le_dev_opt$`Land-locked Developing Countries (LLDC)` %>%
  forecast(h = 5) %>% 
  autoplot()
p5 <- le_dev_opt$`Small Island Developing States (SIDS)` %>%
  forecast(h = 5) %>% 
  autoplot()

# display plots side by side
gridExtra::grid.arrange(p1, p2, p3, p4, p5, ncol = 2)
```


### Create a function to plot with combined data

```{r}
plot_forecast <- function(forecast_object, historical_data) {
  # Get forecasted data as a data frame
  forecast_data <- as.data.frame(forecast(forecast_object, h = 5))
  forecast_time <- seq(max(time(historical_data)) + 1, by = 1, length.out = nrow(forecast_data))
  
  # Combine historical and forecasted data
  combined_data <- data.frame(
    Time = c(time(historical_data), forecast_time),
    Value = c(as.numeric(historical_data), forecast_data$`Point Forecast`),
    Type = rep(c("Historical", "Forecast"), times = c(length(historical_data), nrow(forecast_data)))
  )
  
  # Create data frame for the ribbon layer
  ribbon_data <- data.frame(
    Time = forecast_time,
    ymin = forecast_data$`Lo 80`,
    ymax = forecast_data$`Hi 80`
  )
  
  # Plot the combined data
  ggplot(combined_data, aes(x = Time, y = Value, color = Type)) +
    geom_line() +
    geom_ribbon(data = ribbon_data, aes(x = Time, ymin = ymin, ymax = ymax), fill = "blue", alpha = 0.2) +
    labs(title = deparse(substitute(forecast_object)), y = "Life Expectancy", x = "Time") +
    theme_minimal()
}
```

### Generate and display the plots

```{r}
library(gridExtra)
library(ggplot2)

p1 <- plot_forecast(le_dev_opt$`Least developed countries`, le_dev_ls$`Least developed countries`)
p2 <- plot_forecast(le_dev_opt$`More developed regions`, le_dev_ls$`More developed regions`)
p3 <- plot_forecast(le_dev_opt$`Less developed regions, excluding least developed countries`, le_dev_ls$`Less developed regions, excluding least developed countries`)
p4 <- plot_forecast(le_dev_opt$`Land-locked Developing Countries (LLDC)`, le_dev_ls$`Land-locked Developing Countries (LLDC)`)
p5 <- plot_forecast(le_dev_opt$`Small Island Developing States (SIDS)`, le_dev_ls$`Small Island Developing States (SIDS)`)

# display plots side by side
gridExtra::grid.arrange(p1, p2, p3, p4, p5, ncol = 2)
```

This modified code creates a function `plot_forecast` that combines the historical and forecasted data for smoother transitions. The historical data and forecasted data are then plotted together, ensuring continuity between the two. This should eliminate the visual disconnect you observed.




## ACF and PACF plots

```{r}
# install.packages("astsa")
library(astsa)
acf2(le_dev_ls$`Least developed countries`)
sarima(le_dev_ls$`Least developed countries`, 0, 1, 2)
sarima.for(le_dev_ls$`Least developed countries`, 5, 0, 1, 2)

# acf2(le_dev_ls$`More developed regions`)
```

## ADF test

```{r}
library(tseries)
le_dev <- le_clean %>%
  filter(entity %in% development_status) %>% 
  select(entity, year, diff) %>% drop_na()
adf.test(le_dev %>% filter(entity == "More developed regions") %>% pull(diff))
```


# Global Vaccination Coverage

coverage: one-year-old immunized %

```{r}
vaccination_df <- read_csv("data/global-vaccination-coverage.csv")
vaccination_df %>% glimpse()
vaccination_clean <- janitor::clean_names(vaccination_df)
colnames(vaccination_clean) <- sub("_.*", "", colnames(vaccination_clean))

vaccination_long <- vaccination_clean %>% 
  pivot_longer(cols = -c(entity, code, year), 
               names_to = "vaccine", 
               values_to = "coverage")

```

```{r}
visdat::vis_miss(vaccination_long)
```

```{r}
WHO_regions <- stringr::str_subset(vaccination_df$Entity, "WHO") %>% unique()
p_vac <- vaccination_long %>% 
  filter(entity %in% WHO_regions) %>%
  ggplot(aes(x = year, y = coverage, color = vaccine)) +
  geom_line() + geom_point() +
  facet_wrap(~entity) +
  labs(title = "Global Vaccination Coverage by Year",
       x = "Year",
       y = "Coverage (%)")
ggplotly(p_vac)
```

# Multivariate Time Series Analysis: Life Expectancy and Vaccination Coverage

We focus on the 5 vaccines with coverage rate data since 1980.
- BCG: Tuberculosis
- DTP3: Diphtheria/Tetanus/Pertussis
- MCV1: Measles, first dose
- POL3: Polio
- RCV1: Rubella

```{r}
vaccines <- c("BCG", "DTP3", "MCV1", "POL3", "RCV1")
vaccines_world <- vaccination_clean %>% 
  filter(entity == "World") %>%
  select(bcg, dtp3, mcv1, pol3, rcv1)
```

```{r}
vaccines_world_ts <- ts(vaccines_world[-1], start = min(vaccines_world$year), end = max(vaccines_world$year))
```

```{r}
le_world <- le_clean %>% filter(entity == "World" & year >= 1980) %>% select(year, period_life_expect)
le_world_ts <- ts(le_world$period_life_expect, start = 1980, end = 2021)
```

fit model

```{r}
fit <- auto.arima(le_world_ts, xreg = vaccines_world_ts)
summary(fit)
```

```{r}
# Predict future values
forecast_period <- 5
future_vaccine_cov = matrix(
  c(84, 81, 81, 80, 66,
    85, 83, 83, 82, 68,
    86, 83, 84, 83, 70,
    87, 84, 84, 84, 71,
    88, 84, 84, 84, 72), ncol = 5, byrow = TRUE)
colnames(future_vaccine_cov) <- c("bcg", "dtp3", "mcv1", "pol3", "rcv1")
# Forecast
forecast_fit <- forecast(fit, xreg = future_vaccine_cov, h = forecast_period)

# Plot the forecast
autoplot(forecast_fit)
```

